Effects of bariatric surgery loss on resting-state functional connectivity of the reward and default mode network: a pre-registered analysis

### 2021-03-05 manuscript/draft_rsfmri.Rmd
changes of VW and EM added by FB

Obesity imposes serious health risks and is consistently associated with differences in brain activity, including resting-state functional connectivity. Effective treatment can be achieved by bariatric surgery, yet its effects on functional connectivity of brain networks involved in reward processingand intereoception are still under debate. In this pre-registered analysis, we aimed to determine the effects of bari-atric surgery and weight loss on functional connectivity in the reward network and default mode network (DMN. 
Therefore, 33 bariatric surgery patients (males/females, BMI, age etc) and 15 obese waiting-list control patients (males/females, BMI, age etc) underwent MRI at baseline (before surgery/waiting period), and after 6 and 12 months. We pre-registered (https://osf.io/f8tpn and https://osf.io/59bh7/) the whole-brain time-by-group interaction on network connectivity, assessed between baseline and followup and for two denoising pipelines. We also pre-registered a time-by-group interaction analysis of aggregated FC within networks.  In exploratory analyses, we investigated all three time points and assessed the effects of weight loss and head motion. 
According to whole-brain marginal models, bariatric surgery compared to waiting did not significantly affect functional connectivity of the the reward and DMN networks (FWE p > 0.05). There also was no interaction effect on within network FC in pre-registered analysis of aggregated FC. Exploratory analyses also did not show an interaction effect for three time points, but when adding differences in head motion within and between individuals using mean framewise displacement, surgery-related BMI decrease and higher average head motion resulted in stronger functional connectivity in medial posterior frontal regions within the reward network. 
This pre-registered analysis of longitudinal neuroimaging data did not support a strong effect of bariatric surgery, compared to waiting, on reward and DMN network functional connectivity after 6 months in obese patients. Exploratory analyses indicated that differences in head motion might have confounded effects of weight loss on reward-related functional connectivity, yet definite conclusions remain difficult. Future studies with a more rigorous control of within-scanner head motion during data acquisition are warranted to further understand the effects of obesity surgery on functional connectivity measures.

@@

Precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated. Resting-state functional magnetic resonance imaging (rsrsfMRI) is a technique to investigate the dynamic organization of the brain. Functional connectivity (FC) networks, i.e. brain regions with correlated neural activity over time, are in anatomical correspondence with specific brain networks involved in cognitive processes, including attention and executive control [@Smith_2009]. The reward network (RN), processing hedonic value and internal motivation, and the default mode network (DMN), a higher-order network, involved in interoception and governing shifts between external-internal processes,  are promising candidates to mediate altered central regulation of food intake after bariatric surgery.
The reward network (RN), comprises the ventromedial prefrontal cortex (vmPFC), the striatum (STR), the amygdala (AMY) and the anterior insula (antINS) [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of reward network regions during anticipation of (high-caloric) food cues, and in contrast, reduced activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. resting-state rsfMRI studies also shown increased local FC of re-ward network regions, i.e. NAcc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. In bariatric surgery patients, connectivity  within the RN (e.g. putamen, OFC) might be normalized by the surgery, however the evidence is limited due to a lack of longitudinal obese control groups  [@Duan_2020; @Wiemerslage_2016]. Possibly, a reconfiguration of the fronto-striatal brain networks could emerge from altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018] .. 
 The default mode network includes the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015]. Higher BMI and obesity have been associated with a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks in several resting-state and task-based rsfMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015; @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018]. Studies in bariatric surgery patients have pointed to a normalization of the connectivity between  DMN and cognitive control and salience brain regions after the procedure, yet no study has included a longitudinal control group [@Li_2018; @Olivo_2017; @Frank_2013].
In sum, while there is some evidence hinting to a role for DMN and RN connectivity in altered regulation of food intake after bariatric surgery, the existing evidence is inconclusive. Most studies have investigated small cohorts of patients, without adequate obese control groups, and did not rigorously separate confirmatory from exploratory analyses [@George_2016].
Most resting-state rsfMRI studies in obesity have reported an association of higher BMI and higher head motion during MRI [@Beyer_2017;@Hodgson_2016]. Head motion is a major confounder in resting-state rsfMRI as it increases short-range and disruptes long-range connectivity FC. In this sample, we previously reported a group-by-time interaction on head motion [@Beyer_2020]. Here, we thus aimed to rigorously control for motion-related variance. 
In this pre-registered study, our hypotheses were:
1. Whole-brain FC of the RN and DMN changes differently from baseline to followup in the bariatric surgery compared to waiting-list control group
2. Within-network FC of the RN and DMN changes differently from baseline to followup in the bariatric surgery compared to waiting-list control group

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study investigated the effects of bariatric surgery on brain structure and function in a prospective design at the Charité University Medicine Berlin, Germany. For more details see Prehn et al. (2020). We used all data acquired until April 2019.
The study protocol was in accordance with the Helsinki Declaration and approved by the ethics committee of the Charité University Medicine Berlin. The study was registered at clinicaltrials.gov as NCT01554228. We preregistered the present resting-state analyses on the OSF https://osf.io/yp42s. We made additional changes https://osf.io/59bh7/ to the pre-registration after preprocessing as we realized some aspects of the analysis were inadequately described in the initial pre-registration.
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA -->

Participants were recruited from the Center for Bariatric and Metabolic Surgery at the Charité University Medicine Berlin. Inclusion criteria were, in accordance with German guidelines for bariatric surgery, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013]. Participants were aged between $18$ and $70$ years and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score < $34$).
 In total, $51$ participants out of the originally enrolled $69$ subjects received MRI, and we analyzed a total of $101$ data points. Five data points of three subjects had to be excluded due to bad anatomical image quality (see below).  
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`). Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery/baseline appointment to capture both phases of rapid weight-loss and maintenance [@Maciejewski_2016]. Analyses were performed on all participants that provided at least one data point of rsfMRI data. In more detail, `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for one time point.

@@

Patients arrived in the morning (between 07:00 and 12:00 a.m.) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one hour break for breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020)).

## RsrsfMRI acquisition

@@

## Denoising pipelines
Among several denoising pipelines, ICA-AROMA provides an acceptable trade-off between the mitigation of motion-FC correlation and the introduction of a distance-dependence on motion-FC relation-ship [@Parkes_2018]. [@Ciric_2017] showed that GSR is very efficient in re-moving the correlation of head motion and FC, and outperforms ICA-AROMA alone. Yet, it introduces spatial dependency and spurious correlations. Thus, we pre-registered the main analyses to be performed with these two denoising pipelines.

### ICA-AROMA preprocessing   
As described in [@pruim2015ica], the minimally preprocessed data underwent independent component analysis (ICA) using FSL's `melodic` and nuisance components were classified according to four spectral and spatial criteria. Finally, the timeseries of the  nuisance components were regressed from the raw rsfMRI timeseries non-aggressively.
As recommended in the original paper, we performed regression of WM and CSF signal on the ICA-AROMA result. We used CompCor to estimate 5 variance components from a combined WM and CSF mask, which was further eroded using `fslmaths -nan -thr 0.99 -ero -bin` [@Behzadi_2007;@Muschelli_2014]. After regression of these nuisance components from the data, we performed high-pass filtering at 0.01 Hz.

@@

### Functional connectivity (FC)
To derive RN and DMN FC maps, we used nucleus accumbens (NAcc) and precu-neus as seed regions of interest (ROI), respectively. We did not select vmPFC for the RN due to low SNR. Because we knew about the large amount of motion in this dataset beforehand, we decided against ICA to construct the networks, as ICA is recommended to work on minimally preprocessed data. Based on FreeSurfer’s aseg.mgz and Desikan-Killiany parcellation, we created seed masks using mri_binarize (thresholded for NAcc at 26,58; precuneus at 1025,2025) and averaged them over hemispheres. Then, we used NiftiLa-belsMasker and NiftiMasker to extract the standardized timeseries from the seed regions and the whole brain. We calculated the Pearson’s correlation between them with numpy.dot, performed r-to-z Fisher-transformation and saved the resulting correlation maps for each processing mode (minimally preprocessed, aro-ma+cc, aroma+gsr). Finally, the connectivity maps were transformed into MNI space using the affine transformation and non-linear warp derived with ANTS dur-ing anatomical preprocessing.

@@

#### Resting-state fMRI quality control
Resting-state rsfMRI quality control was performed according to the protocol by @Ciric_2018. 
Head motion is a major confounder in rsfMRI. We calculated mean framewise displacement (mFD) from the outputs of `MCFLIRT` according to @Power_2012.
As pre-registered, we did not exclude anybody based on high average head motion [@Beyer_2020]. Instead, we applied best-practice motion correction (AROMA+CC and AROMA+CC+GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018].

**Individual level resting-state rsfMRI QC**
```{r carpetplot, out.width="50%", out.height="30%", fig.cap="Carpet plot for quality control", fig.subcap = c("Example 1 with strong amount of structured noise", "Example 2 with less amount of structured noise"), fig.show='hold',fig.align='center', echo=FALSE}
carpets=c("../report/carpetplotadi006fu2.jpeg","../report/carpetplotadi16fu.jpeg")
knitr::include_graphics(carpets)
```
On the individual level, we used `plot_carpet` from `niworkflows.viz.plots` to generate carpet plots which depict denoising quality. Figure \@ref(fig:carpetplot) shows an exemplary carpet plot with (from top to bottom): percent outliers defined by AFNI, mFD, `OutlierCount`, DVARS (spatial root mean square of the data after temporal differencing),voxelwise timeseries from minimally preprocessed, AROMA non-aggressive and aggressive denoising, CC and CC + GSR data (red: GM voxels, green: WM voxels, blue: ventricle, green: cerebellum). All carpet plots were visually checked and if there was still structured noise in the CC+GSR-denoised functional data (such as black/white stripes, signal dropout), the participant was given a rating of 1 in `QA residuals`.

**Group level resting-state rsfMRI QC**

First, we inspected parameters of head motion (mFD) and a measure of whole-brain signal variation which is reflective of noise in the data (DVARS).
```{r figDvarsmFD, out.width="100%", fig.cap='Summary of quality measures DVARS and mFD', fig.show='hold',fig.align='center', echo=FALSE}
figList <- readRDS(file = "../report/fig/figDvarsmFDList.rds")
#(figList$fig1 | figList$fig2) / figList$fig3

lay <- rbind(c(1,1,2,2),
             c(NA,3,3,NA))
gridExtra::grid.arrange(grobs = figList, layout_matrix = lay)
```

As described in @Beyer_2020, there was a group-by-time interaction on mFD. The intervention group reduced in head motion in the follow-up assessments compared to the control group.
Further, DVARS did not qualitatively differ between groups and time points, and the correlation of DVARS and mFD remained similar over time points (see Figure \@ref(fig:figDvarsmFD).
Following the protocol by [@Ciric_2018], we evaluated mFD-FC relationships by computing a functional connectome based on [@Power_2012]. In more detail, we used the MNI coordinates of 246 cortical and subcortical spherical seed regions with 5mm radius in `NiftiSpheresMasker`, and calculated the Spearman rank correlation between these time series. Further, we evaluated the distance-dependence of mFD-FC correlations by correlating the euclidean distance between nodes with the mFD-QC correlation and report mean, median and standard deviations of mFD, DVARS and number of outliers over all participants.
```{r mFDFC, echo=FALSE}
res <- read.csv("../report/FDFCcorrelations.csv")
rownames(res) <- c("minimally processed", "AROMA", "AROMA + CC", "AROMA + CC + GSR")
colnames(res) <- c('mean FD-QC', 'median FD-QC', 'sig. vertex', 'sig. vertex BH', 'distance-FD-QC', 'pval')
```
```{r tableQcfc, out.width="100%", echo=FALSE}
tab_Qcfc <-
  knitr::kable(
    round(res, 3),
    col.names = colnames(res),
    row.names = TRUE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "Summary of quality metrics for different denoising pipelines across conditions and time points"
  ) %>%
  column_spec(c(2:6), width = "1.8cm")
tab_Qcfc
```
Table \@ref(tab:tableQcfc) shows that in the minimally preprocessed data, mFD and FC between nodes were significantly positively correlated (mean Spearman's $r$ = `r round(res[1,1],2)`). This relationship was negatively associated with distance between nodes (mean Spearman's $r$ = `r round(res[1,5],2)`).  
Denoising pipelines (AROMA, AROMA+CC, AROMA+CC+GSR) gradually reduced the correlation of mFD and connectivity, and reduced the distance dependency. Contrary to previous reports, GSR did not exercabate mFD-FC distance dependency [@Power_2015]. Overall, AROMA+CC+GSR performed best, with almost no mFD-FC correlation (mean Spearman's $r$ = `r round(res[4,1],2)`), and minimal positive distance dependence (mean Spearman's $r$ = `r round(res[4,5],2)`)
In addition, for models showing a significant time-by-group interaction, we planned to perform a sensitivity analysis excluding the 10% data sets with highest mFD.

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and reward network over all participants and NAccs, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with $N = 1000$. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average connectivity from these masks for GSR regressed data.

@@

<!-- MODEL SETUP --> 
#### Confirmatory Analysis
We tested the preregistered hypothesis of a  time by group interaction:  $FC=group + time + group * time$. 
As preregistered, we performed the analysis for baseline and followup time points only, in both denoising pipelines and adjusting for no confounders (CA 1), age, sex and average mFD of both timepoints (CA2) and age,sex, average mFD and baseline BMI (CA3). 
 In this model, time was represented as factor because we took into account possible non-linear time courses in the increase and decrease of FC over the course of one year, which may occur depending on the phase of weight management [@Olivo_2017]. The resulting factorial design contained one regressor for each time point per group (see supplements for depiction of design matrix).

#### Exploratory analysis
##### Group-by-time interaction in three time points
Then, we extended our whole-brain analysis of the group-by-time point interaction to the data from all three timepoints. This analysis had not been preregistered. Here, we used individual mFD values (not averaged) as covariate to capture variance in mFD change over time points.We investigated two models (EA1.1: group x time, age, sex and EA1.2: group x time, age, sex and mFD).

##### Average and change BMI
We further  explored the relationship of average and change in BMI in a separate group of models. As proposed by @Guillaume_2014, we calculated the between- and within subject centered values of BMI. This model, containing average BMI and BMI variability, allowed us to disentangle the differential effects of overall BMI and change in BMI on FC. We first estimated their effects in a model adjusting for baseline age and sex (Model EA2.1) and then additionally controlling for log mean mFD (Model EA 2.2). As we previously reported correlated change in BMI and mFD in this sample [@Beyer_2019], we explored a refined model including average BMI and logmFD and change in both measures, along with baseline age and sex (Model EA 2.3c). Here, we aimed to see whether any effect of change in BMI would be detectable when adjusting for the change in head motion. 
For a better understanding of the unique contribution of average and longitudinal change FD measures, we separately employed a third post-hoc, exploratory model (EA 2.4), solely including the head motion measures as covariates of interest: $FC = between-subject FD + within-subject FD$ age and sex as nuisance covariates.

@@

All code is available in the github repository [https://github.com/hsx1/adi2_rsrsfMRI](https://github.com/hsx1/adi2_rsrsfMRI). The code includes script for data preparation and analysis of different models.

>>>>

Obesity imposes serious health risks and is consistently associated with differences in brain activity, including resting-state functional connectivity. Effective treatment can be achieved by bariatric surgery, yet its effects on functional connectivity of brain networks involved in reward processingand intereoception are still under debate. In this pre-registered analysis, we aimed to determine the effects of bari-atric surgery and weight loss on functional connectivity in the RN and default mode network (DMN. 
Therefore, 33 bariatric surgery patients (males/females, BMI, age etc) and 15 obese waiting-list control patients (males/females, BMI, age etc) underwent MRI at baseline (before surgery/waiting period), and after 6 and 12 months. We pre-registered (https://osf.io/f8tpn and https://osf.io/59bh7/) the whole-brain time-by-group interaction on network connectivity, assessed between baseline and followup and for two denoising pipelines. We also pre-registered a time-by-group interaction analysis of aggregated FC within networks.  In exploratory analyses, we investigated all three time points and assessed the effects of weight loss and head motion. 
According to whole-brain marginal models, bariatric surgery compared to waiting did not significantly affect functional connectivity of the the reward and DMN networks (FWE p > 0.05). There also was no interaction effect on within network FC in pre-registered analysis of aggregated FC. Exploratory analyses also did not show an interaction effect for three time points, but when adding differences in head motion within and between individuals using mean framewise displacement, surgery-related BMI decrease and higher average head motion resulted in stronger functional connectivity in medial posterior frontal regions within the RN. 
This pre-registered analysis of longitudinal neuroimaging data did not support a strong effect of bariatric surgery, compared to waiting, on reward and DMN network functional connectivity after 6 months in obese patients. Exploratory analyses indicated that differences in head motion might have confounded effects of weight loss on reward-related functional connectivity, yet definite conclusions remain difficult. Future studies with a more rigorous control of within-scanner head motion during data acquisition are warranted to further understand the effects of obesity surgery on functional connectivity measures.

@@

Precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated. Resting-state functional magnetic resonance imaging (rsfMRI) is a technique to investigate the dynamic organization of the brain. Functional connectivity (FC) networks, i.e. brain regions with correlated neural activity over time, are in anatomical correspondence with specific brain networks involved in cognitive processes, including attention and executive control [@Smith_2009]. The RN (RN), processing hedonic value and internal motivation, and the default mode network (DMN), a higher-order network, involved in interoception and governing shifts between external-internal processes,  are promising candidates to mediate altered central regulation of food intake after bariatric surgery.
The RN (RN), comprises the ventromedial prefrontal cortex (vmPFC), the striatum (STR), the amygdala (AMY) and the anterior insula (antINS) [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of RN regions during anticipation of (high-caloric) food cues, and in contrast, reduced activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. resting-state rsfMRI studies also shown increased local FC of re-ward network regions, i.e. NAcc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. In bariatric surgery patients, connectivity  within the RN (e.g. putamen, OFC) might be normalized by the surgery, however the evidence is limited due to a lack of longitudinal obese control groups  [@Duan_2020; @Wiemerslage_2016]. Possibly, a reconfiguration of the fronto-striatal brain networks could emerge from altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018] . 
 The DMN includes the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex (approx. BA 39) [@Raichle_2015]. Higher BMI and obesity have been associated with a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks in several resting-state and task-based rsfMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015; @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018]. Studies in bariatric surgery patients have pointed to a normalization of the connectivity between DMN and cognitive control and salience brain regions after the procedure, yet no study has included a longitudinal control group [@Li_2018; @Olivo_2017; @Frank_2013].
In sum, while there is some evidence hinting to a role for DMN and RN FC in altered regulation of food intake after bariatric surgery, the existing evidence is inconclusive. Most studies have investigated small cohorts of patients, without adequate obese control groups, and did not rigorously separate confirmatory from exploratory analyses [@George_2016].
Most rsfMRI studies in obesity have reported an association of higher BMI and higher head motion during MRI [@Beyer_2017;@Hodgson_2016]. Head motion is a major confounder in rsfMRI as it increases short-range and disrupts long-range connectivity FC. In this sample, we previously reported a group-by-time interaction on head motion [@Beyer_2020]. Here, we thus aimed to rigorously control for motion-related variance. 
In this pre-registered study, our confirmatory hypotheses were:
1. Whole-brain FC of the RN and DMN changes differently from baseline to followup in the bariatric surgery compared to a waiting-list control group
2. Within-network FC of the RN and DMN changes differently from baseline to followup in the bariatric surgery compared to a waiting-list control group

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study investigated the effects of bariatric surgery on brain structure and function in a prospective design at the Charité University Medicine Berlin, Germany. For more details see [@Prehn_2020]. We used all data acquired until April 2019.
The study protocol was in accordance with the Helsinki Declaration and approved by the ethics committee of the Charité University Medicine Berlin. The study was registered at clinicaltrials.gov as NCT01554228. We pre-registered the present resting-state analyses on the OSF https://osf.io/yp42s. We made additional changes https://osf.io/59bh7/ to the pre-registration after preprocessing as we realized some aspects of the analysis were inadequately described in the initial pre-registration.
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA -->

Participants were recruited from the Center for Bariatric and Metabolic Surgery at the Charité University Medicine Berlin. Inclusion criteria were, in accordance with German guidelines for bariatric surgery, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013]. Participants were aged between $18$ and $70$ years and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score < $34$).
 In total, $51$ participants out of the originally enrolled $69$ subjects received MRI, and we analyzed a total of $101$ data points. Five data points of three subjects had to be excluded due to bad anatomical image quality (see below).  
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`). Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery/baseline appointment to capture both phases of rapid weight-loss and maintenance [@Maciejewski_2016]. 
Analyses were performed on all participants that provided at least one data point of rsfMRI data. `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for one time point (for more details see Table S1). The pre-planned analysis of changes from baseline to follow-up included $24$ participants with both time points, in total $72$ data points.

@@

Participants arrived in the morning (between 07:00 and 12:00 a.m.) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one hour break for breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020)).

## MRI acquisition

@@

## Denoising pipelines
Among several denoising pipelines, ICA-AROMA provides an acceptable trade-off between the mitigation of motion-FC correlation and the introduction of a distance-dependence on motion-FC relation-ship [@Parkes_2018]. @Ciric_2017 showed that GSR is very efficient in re-moving the correlation of head motion and FC, and outperforms ICA-AROMA alone. Yet, it introduces spatial dependency and spurious correlations. Thus, we pre-registered the main analyses to be performed with these two denoising pipelines.

### ICA-AROMA denoising pipeline   
As described in @pruim2015ica, the minimally preprocessed data underwent independent component analysis (ICA) using FSL's `melodic` and nuisance components were classified according to four spectral and spatial criteria. Finally, the timeseries of the  nuisance components were regressed from the raw rsfMRI timeseries non-aggressively.
As recommended in the original paper, we performed regression of WM and CSF signal on the ICA-AROMA result. We used CompCor to estimate 5 variance components from a combined WM and CSF mask, which was further eroded using `fslmaths -nan -thr 0.99 -ero -bin` [@Behzadi_2007;@Muschelli_2014]. After regression of these nuisance components from the data, we performed high-pass filtering at 0.01 Hz.

@@

### Functional connectivity (FC)
To derive RN and DMN FC maps, we used nucleus accumbens (NAcc) and precu-neus as seed regions of interest (ROI), respectively. We did not select vmPFC for the RN due to low SNR. Because we knew about the large amount of motion in this dataset beforehand, we decided against ICA to construct the networks, as ICA is recommended to work on minimally preprocessed data. Based on FreeSurfer’s aseg.mgz and Desikan-Killiany parcellation, we created seed masks using mri_binarize (thresholded for NAcc at 26,58; precuneus at 1025,2025) and averaged them over hemispheres. Then, we used NiftiLa-belsMasker and NiftiMasker to extract the standardized timeseries from the seed regions and the whole brain. We calculated the Pearson’s correlation between them with numpy.dot, performed r-to-z Fisher-transformation and saved the resulting correlation maps for each processing mode (minimally preprocessed, aro-ma+cc, aroma+gsr). Finally, the connectivity maps were transformed into MNI space using the affine transformation and non-linear warp derived with ANTS during anatomical preprocessing.

@@

#### RsfMRI quality control
RsfMRI quality control was performed according to the protocol by @Ciric_2018. 
Head motion is a major confounder in rsfMRI. We calculated mean framewise displacement (mFD) according to @Power_2012, and log-transformed it for further analysis (logmFD).
As pre-registered, we did not exclude anybody based on high average head motion [@Beyer_2020]. Instead, we applied best-practice motion correction (AROMA+CC and AROMA+CC+GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018].

##### Individual level resting-state rsfMRI QC
```{r carpetplot, out.width="50%", out.height="30%", fig.cap="Carpet plot for quality control", fig.subcap = c("Example 1 with strong amount of structured noise", "Example 2 with less amount of structured noise"), fig.show='hold',fig.align='center', echo=FALSE}
carpets=c("../report/carpetplotadi006fu2.jpeg","../report/carpetplotadi16fu.jpeg")
knitr::include_graphics(carpets)
```
On the individual level, we used `plot_carpet` from `niworkflows.viz.plots` to generate carpet plots which depict denoising quality. Figure \@ref(fig:carpetplot) shows an exemplary carpet plot with (from top to bottom): percent outliers defined by AFNI, mFD, `OutlierCount`, DVARS (spatial root mean square of the data after temporal differencing),voxelwise timeseries from minimally preprocessed, AROMA non-aggressive and aggressive denoising, CC and CC + GSR data (red: GM voxels, green: WM voxels, blue: ventricle, green: cerebellum). All carpet plots were visually checked and if there was still structured noise in the CC+GSR-denoised functional data (such as black/white stripes, signal dropout), the participant was given a rating of 1 in `QA residuals`, and checked again visually. No participant was excluded based on this visual inspection.

##### Group level resting-state rsfMRI QC

As described in @Beyer_2020, there was a group-by-time interaction on mFD (see Figure \@ref(fig:figDvarsmFD), upper right panel).  The intervention group reduced in head motion in the follow-up assessments compared to the control group. 
For this study, we further explored mFD and its relation to DVARS, the root mean square of the temporal change of the voxel-wise signal at each time point.
```{r figDvarsmFD, out.width="100%", fig.cap='Summary of quality measures DVARS and mFD', fig.show='hold',fig.align='center', echo=FALSE}
figList <- readRDS(file = "../report/fig/figDvarsmFDList.rds")
#(figList$fig1 | figList$fig2) / figList$fig3

lay <- rbind(c(1,1,2,2),
             c(NA,3,3,NA))
gridExtra::grid.arrange(grobs = figList, layout_matrix = lay)
```
DVARS did not qualitatively differ between groups and time points, and the correlation of DVARS and mFD remained similar over time points (see Figure \@ref(fig:figDvarsmFD).
To further assess the confounding of head motion and FC, we evaluated mFD-FC relationships by computing a functional connectome based on the protocol by [@Ciric_2018]. We used the MNI coordinates of 246 cortical and subcortical spherical seed regions from @Power_2012 with 5mm radius in `NiftiSpheresMasker`, and calculated the Spearman rank correlation between these time series. Then, we calculated the distance-dependence of mFD-FC correlations by correlating the euclidean distance between nodes with the mFD-QC correlation.
```{r mFDFC, echo=FALSE}
res <- read.csv("../report/FDFCcorrelations.csv")
rownames(res) <- c("minimally processed", "AROMA", "AROMA + CC", "AROMA + CC + GSR")
colnames(res) <- c('mean FD-QC', 'median FD-QC', 'sig. vertex', 'sig. vertex BH', 'distance-FD-QC', 'pval')
```
```{r tableQcfc, out.width="100%", echo=FALSE}
tab_Qcfc <-
  knitr::kable(
    round(res, 3),
    col.names = colnames(res),
    row.names = TRUE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "Summary of quality metrics for different denoising pipelines across conditions and time points"
  ) %>%
  column_spec(c(2:6), width = "1.8cm")
tab_Qcfc
```
In the minimally preprocessed data, mFD and FC between nodes were significantly positively correlated (mean Spearman's $r$ = `r round(res[1,1],2)`)(see Table \@ref(tab:tableQcfc)). This relationship was negatively associated with distance between nodes (mean Spearman's $r$ = `r round(res[1,5],2)`).  
Denoising pipelines (AROMA, AROMA+CC, AROMA+CC+GSR) gradually reduced the correlation of mFD and FC, and reduced the distance dependency. Contrary to previous reports, GSR did not exercabate mFD-FC distance dependency [@Power_2015]. Overall, AROMA+CC+GSR performed best, with almost no mFD-FC correlation (mean Spearman's $r$ = `r round(res[4,1],2)`), and minimal positive distance dependence (mean Spearman's $r$ = `r round(res[4,5],2)`)
In addition, for models showing a significant time-by-group interaction, we planned to perform a sensitivity analysis excluding the 10% data sets with highest mFD.

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and RN over all participants and NAccs, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with $N = 1000$. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average FC from these masks for GSR-data.

@@

<!-- MODEL SETUP --> 
#### Confirmatory Analysis
We tested the pre-registered hypothesis of a  time by group interaction:  $FC=group + time + group * time$. 
As pre-registered, we performed the analysis for baseline and followup time points only, in both denoising pipelines and adjusting for no confounders (CA 1), age, sex and average logmFD of both timepoints (CA2) and age,sex, average logmFD and baseline BMI (CA3). 
 In this model, time was represented as factor because we took into account possible non-linear time courses in the increase and decrease of FC over the course of one year, which may occur depending on the phase of weight management [@Olivo_2017]. The resulting factorial design contained one regressor for each time point per group (see supplements for depiction of design matrix).

#### Exploratory analysis
##### Group-by-time interaction in three time points
Then, we extended our whole-brain analysis of the group-by-time point interaction to the data from all three timepoints. This analysis had not been pre-registered. Here, we used individual logmFD values (not averaged) as covariate to capture variance in logmFD change over time points.We investigated two models (EA1.1: group x time, age, sex and EA1.2: group x time, age, sex and logmFD).

##### Average and change BMI
We further  explored the relationship of average and change in BMI in a separate group of models. As proposed by @Guillaume_2014, we calculated the between- and within subject centered values of BMI. This model, containing average BMI and BMI variability, allowed us to disentangle the differential effects of overall BMI and change in BMI on FC. We first estimated their effects in a model adjusting for baseline age and sex (Model EA2.1) and then additionally controlling for log mFD (Model EA 2.2). As we previously reported correlated change in BMI and mFD in this sample [@Beyer_2019], we explored a refined model including average BMI and logmFD and change in both measures, along with baseline age and sex (Model EA 2.3c). Here, we aimed to see whether any effect of change in BMI would be detectable when adjusting for the change in head motion. 
For a better understanding of the unique contribution of average and longitudinal change FD measures, we separately employed a third post-hoc, exploratory model (EA 2.4), solely including the head motion measures as covariates of interest: $FC = between-subject logmFD + within-subject logmFD$ age and sex as nuisance covariates.

@@

All code is available in the github repository [https://github.com/hsx1/adi2_rsfMRI](https://github.com/hsx1/adi2_rsfMRI). The code includes script for data preparation and analysis of different models.
